\textbf{Advection versus diffusion in brain intraventricular transport}

### 2025-03-27 paper/main.tex
final fig6, update plot codes

\title{\textbf{Advection versus diffusion in intraventricular transport}}
%
%
% Author: Please give full first and last names for authors and
% include * after the name of all corresponding authors

\author[1,*]{Halvor Herlyng}
\author[1]{Ada J.~Ellingsrud}
\author[1]{Miroslav Kuchta} 
\author[2]{Inyoung Jeong}
\author[1,3]{Marie E.~Rognes}
\author[2,*]{Nathalie Jurisch-Yaksi}
\affil[1]{Department of Numerical Analysis and Scientific Computing, Simula Research Laboratory, Oslo, Norway} % Kristian Augusts gate 23, 0164
\affil[2]{Department of Clinical and Molecular Medicine, Norwegian University of Science and Technology, Trondheim, Norway} % Erling Skjalgsons gate 1, 7491
\affil[3]{K.~G.~Jebsen Center for Brain Fluid Research, Oslo, Norway}
\affil[*]{\href{mailto:hherlyng@simula.no}{hherlyng@simula.no}, \href{mailto:nathalie.jurisch-yaksi@ntnu.no}{nathalie.jurisch-yaksi@ntnu.no}}
\date{\vspace{-2em}}

@@

% Abstract should be written in the present tense and impersonal style (i.e., avoid we), and be at most 200 words long
\begin{abstract}
Cerebrospinal fluid (CSF) is integral to brain function, as the fluid
provides mechanical support for the brain and the flow of CSF helps
distribute nutrients, neurotransmitters and metabolites throughout
the central nervous system. CSF flow is partly driven
by beating of motile cilia, hair-like structures that line the ependymal cell layer
of the walls of the brain ventricles. Despite the physiological importance of CSF flow,
the underlying mechanisms of cilia-mediated flow and transport
in the brain ventricles remain yet to be comprehensively resolved.
In this study, we analyze and evaluate the role of motile cilia
in intraventricular CSF flow and transport. We use the geometry
of the zebrafish larval brain ventricles, for which we have detailed knowledge of cilia properties and CSF motion velocities
and frequencies, to develop finite element methods that model flow and transport.
The computational model is validated by in vivo experiments that monitor transport
of a photoconvertible protein secreted in the brain ventricles.
Our results show that while cilia contribute to advection of large particles,
diffusion plays a significant role in the transport of small molecules such as proteins.
We also found that cilia location and the geometry of the ventricular system impact solute distribution.
Altogether, our work presents a computational framework that can be applied to other
ventricular systems, together with new concepts of how molecules are transported within the brain and its ventricles.

@@

% I2: What do we know about the problem and the topic?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia, CSF secretion by the choroid plexus, and pulsatile pressure gradients generated by the
cardiac cycle, the respiration, neural activity, and bodily
movements~\cite{Brodal2016, del2010ependymal, Ringers2020Role, Vinje2019RespiratoryMeasurements,
Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment,
macaulay2022cerebrospinal, mestre2018flow}.
Cilia are hair-like structures found in a vast diversity of
biological organisms and in several human organs,
of which the brain is one of them~\cite{Fulford1986Muco-ciliaryLung,
Pacherres2022CiliaryProduction, Jahn1972LocomotionProtozoa,
Brennen1977FluidFlagella, Ringers2020Role, spassky2013motile,
DGama2025MotileBrain, mitchell2007evolution,
Tsukita2012CoordinatedFeet, BLAKE1974MechanicsMotion,
Reiten2017Motile-Cilia-MediatedComputations, Thouvenin2020OriginCanal}. 
The ventricular surface of the brain, known as the ependyma, is lined by
motile cilia that beat in a coordinated pattern at frequencies of
10--40 Hz to mediate fluid flow~\cite{mitchell2007evolution,
Ringers2020Role, spassky2013motile, Ringers2023NovelEpithelia, roth2025structure}. 
Given their microscopic size, ranging from
5 to 15 microns, cilia contribute primarily to flow near the
ependyma in the human brain~\cite{Siyahhan2014FlowVentricles, Olstad2019CiliaryDevelopment,
Ringers2020Role, Faubel2016Cilia-basedVentricles}. In contrast, cardiac,
respiratory and delta waves at lower frequencies induce pulsatile flow
at the larger scales in the brain ventricular system, subarachnoid space, and
subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
Causemann2025, fultz2019coupled, mestre2018flow, daversin2020mechanisms}.

@@

To validate the model predictions, we performed photoconversion
experiments in transgenic zebrafish larvae expressing the secreted
Dendra2 in their brain ventricles. The approach is inspired by earlier work
where a photoconvertible protein was injected into the ventricle~\cite{fame2016directional}.
By using a transgenic system where the protein is directly secreted into the CSF,
we avoid detrimental effects of intraventricular injection on CSF properties. 
Our photoconversion protocol
consisted of the acquisition of a baseline fluorescence intensity
$F_0$, before localized exposure to a UV laser (at ROI 1) that
converts Dendra2 from a green- to a red-emitting fluorescent
protein. To quantify the transport of photoconverted proteins, we
measured and calculated the change in fluorescence intensity $\Delta F
= (F(t)-F_0)/F_0$ over time (\Crefpanels{fig:fig3}{a}).

@@

The ciliated boundaries in our CSF flow model are regionally homogeneous,
whereas in reality, cilia populate the ventricle walls in a
heterogeneous manner~\cite{Olstad2019CiliaryDevelopment}. Heterogeneous
arrangement of cilia, both in location, beating direction and beating frequency,
has been shown to enhance particle clearance in airway cilia arrays
in mice~\cite{Ramirez-SanJuan2020Multi-scaleArrays}.
Incorporating spatial arrangement heterogeneity in our model could possibly
result in different transport timescales,
but we would not expect notable changes to the general trends of the macroscopic
transport that we study here, since the effects of introducing heterogeneous spacing
of cilia would alter the magnitude of cilia forces, but not their direction.
On the other hand, introducing more heterogeneous beating directionality could
alter the transport timescales as well as the directionality.
In our model, all of the applied cilia
tangential traction is aligned with the rostrocaudal axis, and more heterogeneity
in this alignment to reflect heterogeneity in the beating directionality of
cilia would presumably affect both magnitude and directionality of advective transport.
Previous studies have demonstrated that changes to beating directionality affects transport,
and that specific coordination of the ciliary beating is important in the brain ventricles for
flow and signaling processes~\cite{Eichele2020Cilia-drivenVentricle, Olstad2019CiliaryDevelopment,
Faubel2016Cilia-basedVentricles, Guirao2010CouplingCilia, Afzelius2004CiliaRelatedDiseases},
but also in other phenomena such as maintaining healthy functioning of mucous transport in
airways~\cite{Rayner1996CiliarySyndrome, Schneiter2021Multi-scaleFunction,
Bustamante-Marin2019LackClearance, Tsukita2012CoordinatedFeet} and in nodal flow which establishes
left-right symmetry in mammals~\cite{Yoshiba2014RolesSymmetry, Hirokawa2006NodalAsymmetry,
Sawamoto2006NewBrain}.
Although these studies were mostly on spatial scales resolving individual cilia,
we could envision the same effects in our continuum model,
similar to the results of previous numerical studies~\cite{Ramirez-SanJuan2020Multi-scaleArrays,
Thouvenin2020OriginCanal, Yoshida2022EffectVentricles}.

@@

For genotyping, genomic DNA (gDNA) was isolated from clipped fins of anesthetized adult
fish using 100 \textmu L of PCR lysis buffer (containing 1M tris pH 7--9, 0.5 M EDTA,
tritonX-100, and Proteinase K 0.1 mg/ml) overnight at 50°C.
To stop the lysis reaction, the samples were heated to 95°C for 10 minutes
and then centrifuged at 13000 rpm for 2 minutes.
The supernatant containing gDNA was utilized for KASP assays-based analysis.
The gDNAs were diluted (1:2) with water, and 3 \textmu L of the diluted gDNA was
used to perform the KASP assay following the guidelines of the manufacturer (LGC Biosearch Technologies™).
For each sample well, the \fixme{master} mix contained: 5 \textmu L of master mix; 0.14
\textmu L of assay mix; 1.86 \textmu L of RO water.

>>>>

\title{\textbf{Advection versus diffusion in brain intraventricular transport}}
%
%
% Author: Please give full first and last names for authors and
% include * after the name of all corresponding authors

\author[1,*]{Halvor Herlyng}
\author[1]{Ada J.~Ellingsrud}
\author[1]{Miroslav Kuchta} 
\author[2]{Inyoung Jeong}
\author[1,3]{Marie E.~Rognes}
\author[2,*]{Nathalie Jurisch-Yaksi}
\affil[1]{Department of Numerical Analysis and Scientific Computing, Simula Research Laboratory, Oslo, Norway} % Kristian Augusts gate 23, 0164
\affil[2]{Department of Clinical and Molecular Medicine, Norwegian University of Science and Technology, Trondheim, Norway} % Erling Skjalgsons gate 1, 7491
\affil[3]{K.~G.~Jebsen Center for Brain Fluid Research, Oslo, Norway}
\affil[*]{\href{mailto:hherlyng@simula.no}{hherlyng@simula.no}, \href{mailto:nathalie.jurisch-yaksi@ntnu.no}{nathalie.jurisch-yaksi@ntnu.no}}
\date{\vspace{-2em}} % Remove space between affiliations and abstract

@@

% Abstract should be written in the present tense and impersonal style (i.e., avoid we), and be at most 200 words long
\begin{abstract}
Cerebrospinal fluid (CSF) is integral to brain function, as the fluid
provides mechanical support for the brain and the flow of CSF helps
distribute nutrients, neurotransmitters and metabolites throughout
the central nervous system. CSF flow is partly driven
by beating of motile cilia, hair-like structures that line the ependymal cell layer
of the walls of the brain ventricles. Despite the physiological importance of CSF flow,
the underlying mechanisms of cilia-mediated flow and transport
in the brain ventricles remain yet to be comprehensively resolved.
In this study, the role of motile cilia in brain intraventricular CSF flow and transport
is analyzed and evaluated. The geometry of the zebrafish larval brain ventricles,
for which detailed knowledge of cilia properties and CSF motion velocities
and frequencies exists, was used to develop finite element methods that model flow and transport.
The computational model is validated by in vivo experiments that monitor transport
of a photoconvertible protein secreted in the brain ventricles.
The results show that while cilia contribute to advection of large particles,
diffusion plays a significant role in the transport of small molecules such as proteins.
It is also demonstrated that cilia location and the geometry of the ventricular system impact solute distribution.
Altogether, this work presents a computational framework that can be applied to other
ventricular systems, together with new concepts of how molecules are transported within the brain and its ventricles.

@@

% I2: What do we know about the problem and the topic?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia, CSF secretion by the choroid plexus, and pulsatile pressure gradients generated by the
cardiac cycle, the respiration, neural activity, and bodily
movements~\cite{Brodal2016, del2010ependymal, Ringers2020Role, Vinje2019RespiratoryMeasurements,
Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment,
macaulay2022cerebrospinal, mestre2018flow}.
Cilia are hair-like structures found in a vast diversity of
biological organisms and in several human organs,
of which the brain is one of them~\cite{DGama2025MotileBrain,
Fulford1986Muco-ciliaryLung,
Pacherres2022CiliaryProduction, Jahn1972LocomotionProtozoa,
Brennen1977FluidFlagella, Ringers2020Role, spassky2013motile,
Pellicciotta2020Cilia, mitchell2007evolution,
Tsukita2012CoordinatedFeet, BLAKE1974MechanicsMotion,
Reiten2017Motile-Cilia-MediatedComputations, Thouvenin2020OriginCanal}. 
The ventricular surface of the brain, known as the ependyma, is lined by
motile cilia that beat in a coordinated pattern at frequencies of
10--40 Hz to mediate fluid flow~\cite{mitchell2007evolution,
Ringers2020Role, spassky2013motile, Ringers2023NovelEpithelia, roth2025structure}. 
Given their microscopic size, ranging from
5 to 15 microns, cilia contribute primarily to flow near the
ependyma in the human brain~\cite{Siyahhan2014FlowVentricles, Olstad2019CiliaryDevelopment,
Ringers2020Role, Faubel2016Cilia-basedVentricles}. In contrast, cardiac,
respiratory and delta waves at lower frequencies induce pulsatile flow
at the larger scales in the brain ventricular system, subarachnoid space, and
subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
Causemann2025, fultz2019coupled, mestre2018flow, daversin2020mechanisms}.

@@

To validate the model predictions, we performed photoconversion
experiments in transgenic zebrafish larvae expressing the secreted
Dendra2 in their brain ventricles. The approach is inspired by earlier work
where a photoconvertible protein was injected into the ventricle~\cite{fame2016directional}.
By using a transgenic system where the protein is directly secreted into the CSF,
we avoid detrimental effects of intraventricular injection on CSF properties. 
Our photoconversion protocol
consisted of the acquisition of a baseline fluorescence intensity
$F_0$, before localized exposure to a UV laser (at ROI 1) that
converts Dendra2 from a green- to a red-emitting fluorescent
protein. To quantify the transport of photoconverted proteins, we
measured and calculated the change in fluorescence intensity $\Delta F(t)
= (F(t)-F_0)/F_0$ over time (\Crefpanels{fig:fig3}{a}).

@@

The ciliated boundaries in our CSF flow model are regionally homogeneous,
whereas in reality, cilia populate the ventricle walls in a
heterogeneous manner~\cite{Olstad2019CiliaryDevelopment}. Heterogeneous
arrangement of cilia, both in location, beating direction and beating frequency,
has been shown to enhance particle clearance in airway cilia arrays
in mice~\cite{Ramirez-SanJuan2020Multi-scaleArrays}.
Incorporating spatial arrangement heterogeneity in our model could possibly
result in different transport timescales,
but we would not expect notable changes to the general trends of the macroscopic
transport that we study here, since the effects of introducing heterogeneous spacing
of cilia would alter the magnitude of cilia forces, but not their direction.
On the other hand, introducing more heterogeneous beating directionality could
alter the transport timescales as well as the directionality.
In our model, all of the applied cilia
tangential traction is aligned with the rostrocaudal axis, and more heterogeneity
in this alignment to reflect heterogeneity in the beating directionality of
cilia would presumably affect both magnitude and directionality of advective transport.
Previous studies have demonstrated that changes to beating directionality affects transport,
and that specific coordination of the ciliary beating is important in the brain ventricles for
flow and signaling processes~\cite{Eichele2020Cilia-drivenVentricle, Olstad2019CiliaryDevelopment,
Faubel2016Cilia-basedVentricles, Guirao2010CouplingCilia, Afzelius2004CiliaRelatedDiseases,
Pellicciotta2020Cilia},
but also in other phenomena such as maintaining healthy functioning of mucous transport in
airways~\cite{Rayner1996CiliarySyndrome, Schneiter2021Multi-scaleFunction,
Bustamante-Marin2019LackClearance, Tsukita2012CoordinatedFeet} and in nodal flow which establishes
left-right symmetry in mammals~\cite{Yoshiba2014RolesSymmetry, Hirokawa2006NodalAsymmetry,
Sawamoto2006NewBrain}.
Although these studies were mostly on spatial scales resolving individual cilia,
we could envision the same effects in our continuum model,
similar to the results of previous numerical studies~\cite{Ramirez-SanJuan2020Multi-scaleArrays,
Thouvenin2020OriginCanal, Yoshida2022EffectVentricles}.

@@

For genotyping, genomic DNA (gDNA) was isolated from clipped fins of anesthetized adult
fish using 100 \textmu L of PCR lysis buffer (containing 1M tris pH 7--9, 0.5 M EDTA,
tritonX-100, and Proteinase K 0.1 mg/ml) overnight at 50°C.
To stop the lysis reaction, the samples were heated to 95°C for 10 minutes
and then centrifuged at 13000 rpm for 2 minutes.
The supernatant containing gDNA was utilized for KASP assays-based analysis.
The gDNAs were diluted (1:2) with water. 
The KASP assay was run following the guidelines of the
manufacturer (LGC Biosearch Technologies™) with 3 \textmu L gDNA,
5 \textmu L of master mix, 0.14 \textmu L of assay mix
and 1.86 \textmu L of RO water per well on a 96-well plate.
